---
title: "Advanced Classification of Human Mutations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classification_Mutations_advanced}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Simpati advanced workflow

Advanced example of Simpati workflow on somatic mutation data of cancer patients
This workflow is recommended for who already went through the simplified workflow and is willing
to learn how to work with Simpati more in details.

Advanced because:
- Every operation is made explicit
- Long workflow

Requirements:
- Average R skills
- Reading Simpati article

**Let's clean and prepare the enviroment of the workflow**
We remove every variables, clean the RAM memory and load Simpati library
We set the random seed in order to get always the same results out of this workflow
We set the number of cores in order to run Simpati in parallel

```{r enviroment setup}
#Clean workspace and memory ----
rm(list=ls())
gc()

#Set working directory----
gps0=getwd()
gps0=paste(gps0,"/%s",sep="")
rootDir=gps0
setwd(gsub("%s","",rootDir))

#Load libraries ----
library(Simpati)

#Set variables ----
#Set seed for reproduce the results
set.seed(0)
#Set TRUE if you are running a introduction vignette to understand how to work with Simpati
test_run=TRUE
#Set the number of cores to use in the workflow
n_cores=5
```

**Let's load the data required to run all the operations**

```{r input data}
#Load and prepare data (e.g. example user data and databases for pathway analysis ) ----

#Get omic-specific patient profiles and their clinical data
geno=tcga_data$LIHC$`LIHC_Mutation-20160128`$assay_df;see(geno)
info=tcga_data$LIHC$`LIHC_Mutation-20160128`$clin_df;see(info)

#Simpati wants the info matrix to be a two column matrix
#patient's names | patient's class (e.g. clinical information)
#Here we select the pathologic_stage of the patient's tumour
info=info[,c("patientID","pathologic_stage")];see(info)

#Set name of the dataset
dataset_name="LIHC"
#Set the semantic type of the disease for the disgnet enrichment
disease_type=tcga_data$LIHC$semantic_type;cat(disease_type)
#Set key words associated to the patient's disease
key_words=tcga_data$LIHC$key_words;cat(key_words)

#Gene interaction network
data(huri_net_l)
net=huri_net_l$net_adj;see(net)

#Pathway list
data(pathways_l);print(pathways_l[1:2])

#Reduce pathways in case you are running a introduction vignette
if(test_run){
  pathways_l=pathways_l[1:1000]
  pathways_l=pathways_l[grep("source-MSIGDB",names(pathways_l))]
}

#human protein atlas for normal cell lines needed for the tissue enrichment
data(db_normal_ATLAS);see(db_normal_ATLAS)

#human protein atlas for cancer cell lines needed for the tumor enrichment
data(db_cancer_ATLAS);see(db_cancer_ATLAS)

#disgnet needed for the disease-feature (aka gene) enrichment
data(db_disgnet);see(db_disgnet)
```

**Let's prepare all the data and variables for Simpati**

We first pass all the input data to the ``` data_preparation ``` function.
This one checks and prepares the data to Simpati workflow.
In case this step does not produce errors, then Simpati will be able to run without further problems.
The result of the ``` data_preparation ``` function is a list of elements extracted from the input data.
From now on, Simpati will need only this list and we will not need to pass the input data other times.

```{r data preparation}
#Data preparation
data_l=data_preparation(geno,info,net,pathways_l,n_cores)
#Clean
rm(geno,info)
```

Similarly to the ``` data_preparation ``` function, there is the ``` set_variables ``` function.
It sets and prepares all the variables required by Simpati.
Plus, it takes extra input data if available. For example, the name of the dataset in study, the type
of the disease associated to the disgnet database, key words associated to the patient's disease, the
number of cores to run the workflow in parallel and the percentage of patients to include in the testing
set of the classification. 
The results of the ``` set_variables ``` function is a list of elements extracted from the input data.

```{r variables set up}
#Generate the object list with all the variables needed by the method
vars_l=set_variables(data_l,dataset_name,disease_type,key_words,n_cores=n_cores,n_LOO = 0.3)
#Clean
rm(disease_type,key_words)
```

Let's start the classification steps.

```{r classification}
#Start to measure the running time
start_time <- Sys.time()

#Propagation
vars_l[["geno_p"]]=rwrProp(net=vars_l$net, geno=vars_l$geno, n_cores=vars_l$n_cores);

#Prepare pathways
vars_l[["pathways_indxs_l"]]=clean_pathways(pathways_l=vars_l$pathways_l, geno=vars_l$geno_p, n_cores=vars_l$n_cores)

#Predicts testing profiles
cat("*Making pathway specific predictions of testing profiles \n")
PSNs_info=predict_pats(vars_l);

#Predict testing profiles and get classification performances
res=get_perfs(PSNs_info)

#Stop running time of classification
end_time <- Sys.time()
running_time=as.numeric(difftime(end_time, start_time, units = "mins")[[1]])

#Save results of classification and pathway analysis
cat("Ended classification in",running_time,"\n")

cat("Starting pathway enrichment of classification results \n")

#Get the updated info dataframe
new_info=get_new_info(res,vars_l)
#Get the pathways used in the classification
pathways_used_l=get_used_pathways(PSNs_info,vars_l)

#Data preparation
data_l=data_preparation(vars_l$orig_geno[,new_info$patientID],new_info,vars_l$net,pathways_used_l)

#Generate the object list with all the variables needed by the method
pr_vars_l=set_variables(data_l,vars_l$dataset_name,vars_l$disease_type,vars_l$key_words,n_cores=vars_l$n_cores)

#Propagation
pr_vars_l[["geno_p"]]=rwrProp(net=pr_vars_l$net,geno=pr_vars_l$geno,n_cores=pr_vars_l$n_cores);

#Prepare pathways
pr_vars_l[["pathways_used_indxs_l"]]=clean_pathways(pathways_l=pr_vars_l$pathways_l,geno=pr_vars_l$geno_p,n_cores=pr_vars_l$n_cores)

#Get signature pathways ----
PSN_data_l=get_sign_pathways(pr_vars_l$geno_p, pr_vars_l$pathways_used_indxs_l, pr_vars_l, n_cores=pr_vars_l$n_cores)  

#Get enrichment ----
gene_sets_l=sapply(PSN_data_l$PSN_comp_l,function(x){x$m_prof_l$row_names})
if(length(gene_sets_l)>1){
  enrXpathway_l=get_pathway_enrich(gene_sets_l,pr_vars_l$key_words,pr_vars_l$disease_type,
                                   db_disgnet,db_cancer_ATLAS,db_normal_ATLAS,
                                   n_cores=3, records_only = T)
}else{enrXpathway_l=list()}

cat("Saving results and workflow data \n")

#Save classification and enrichment results -----
PSN_enr_df=PSN_data_l[["PSN_info_df"]]
enr_PSN_df=enrXpathway_l[["enrXpathway_df"]]
if(!is.null(enr_PSN_df)){
  PSN_enr_df=merge(PSN_enr_df,enr_PSN_df,by="pathway_name",all.x=T)
}

save(res,PSNs_info,running_time,PSN_enr_df,file="Classification_results.rda")
save(vars_l,pr_vars_l,PSN_data_l,enrXpathway_l,file="Enrichment_results.rda")


```
